Split-Apply-Combine with Dynamic Grouping

### 2023-02-21 paper/article.Rnw
wip

@@

\section{Formal descriptions and algorithms}
In this section we give a formal description of the algorithm for aggregation
with dynamic grouping.  We start by giving an algorithm for ordinary
split-apply-combine to show where the algorithm must be generalized to allow
for a collapsing scheme.

\subsection{Split-apply-combine}
In order to analyse a data set group by group we need to specify a data set, a
way to split it into groups, and a function that takes a subset of data and
returns an aggregate. Let us introduce some notation for that.

Let $U$ be a finite set, and let $\phi$ be a function that accepts a subset of
$U$ and returns a value in some domain $X$. Here, $U$ represents a data set and
$\phi$ an aggregating function. We split $U$ into groups using the following
notation. Let $A$ be finite set that has no more elements than $U$, and let $f$
be a function that takes an element of $U$ and returns a value in $A$. We can
think of $A$ as a set of group labels, and $f$ as the function that assigns a
label to each element of $U$. This way, $f$ divides $U$ into non-overlapping
groups. We say that $f:U\to A$ is a \emph{partition} of $U$.

@@

\begin{algorithm}[H]
\caption{Split-Apply-Combine}
\label{alg:sac}
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
\Input{A finite set $U$, an aggregator $\phi: 2^U\to X$, and a partition $f:U\to A$}
\Output{$R$: the value of $\phi$ for every part of $U$ as a set of pairs $(a,x)\in
        A\times X$ }
$R = \{\}$\;
\For{$a\in A$}{
  $d=f^{-1}(\{a\})$\tcp*{get subset of $U$}
  $R = R\cup \{(a,\phi(d))\}$\tcp*{aggregate and add to result}
}
\end{algorithm}

Here, the output is collected in a set $R$ (this set is technically a function
$A\to X$ since it pairs each $a\in A$ with one element of $X$). We also
introduced some new notation: the notation $2^U$ is the power set of $U$ and
$f^{-1}$ stands for the formal \emph{inverse} of $f$. That is: you give it a
set of labels from $A$ and it returns all the elements of $U$ that have any of
those labels. This definition is somewhat too general for the current
algorithm, but it will be necessary when we move to collapsing groups.

Let us see how the elements $U$, $f:U\to A$, and $\phi$ are implemented in
practice. Consider the signature of the R \code{aggregate()} function for data
frames (we skip arguments that are not important for the discussion).
\begin{verbatim}
  aggregate(x, by, FUN)
\end{verbatim}
Here, \code{x} is a data frame where each column is a variable $U$ to be
aggregated.  The parameter \code{by} represents the record-wise label(s) that
are used for grouping. So the function $f\to A$ is implemented by forcing the
user to make sure that the position of each label in \code{by} corresponds to
the row number in the data frame \code{x}. The argument \code{FUN} is the
function $\phi$ that aggregates each subset of \code{x}.  Here is an example of how
a user might call this function from the R prompt.
%
<<>>=
aggregate(iris[1:2], by=iris["Species"], FUN=mean)
@
%
Note that the correspondence in position of the \texttt{Species} label and the
record position is implemented by taking them from the same data frame. The output
also reveals in the first column the set $A$: each row corresponds to a unique
value in \texttt{Species} column.

\subsection{Split-apply-combine with collapsing groups}
\label{sect:saccg}
The goal of the algorithm is to compute a value for each part of a dataset,
possibly using values from a larger subset. The input of the algorithm consists
again of a finite set $U$ and an aggregation function $\phi$ that takes a subset
of $U$ and returns a value in some domain $X$.  Compared to
Algorithm~\ref{alg:sac} we need two extra inputs. First, we need a user-defined
function that checks whether a given subset $d$ of $U$ is suitable for
computing $\phi(d)$.  We will call this function $\beta: 2^U\to \mathbb{B}$,
where $\mathbb{B}=\{\texttt{True},\texttt{False}\}$.  Typical tests are
checking whether there are sufficient records available, or whether certain
variables have a low enough fraction of missing values. Second, we need a
\emph{collapsing scheme}
\begin{equation}
U\xrightarrow{f_0}A_0\xrightarrow{f_1}A_1\xrightarrow{f_2}\cdots\xrightarrow{f_n}A_n.
\label{eq:collapsingsequence}
\end{equation}
A collapsing scheme is a sequence of partitions where each $f_i:A_{i-1}\to A_i$
partitions $A_{i-1}$ into $|A_i|$ groups.

%Collapsing schemes can be represented as tables. For example, the European NACE
%classification is a hierarchical classification of economic activity. Each
%group is encoded as a 4-digit code where the first digit indicates the highest
%grouping level, and the last digit the most detailed level. This induces a
%collapsing sequence by combining groups that have the same parent in the
%hierarchy. Below is an example, based on a small piece of the NACE code list.
%%
%\begin{equation}
%\begin{tabular}{rrr}
%\hline
% $A_0$& $A_1$ & $A_2$\\
%\hline
%0111&   011 & 01\\
%0112&   011 & 01\\
%0113&   011 & 01\\
%0121&   012 & 01\\
%0122&   012 & 01\\
%0123&   012 & 01\\
%0124&   012 & 01\\
%\hline
%\end{tabular}
%\label{eq:tablerep}
%\end{equation}
%
%Here, $A_0$ is the most detailed partition, splitting a dataset into seven
%groups, relating to $\{0111,\ldots,0124\}$. In $A_1$ there are two groups.
%The first, labeled $01.1$ is constructed by combining $0111, 0112$ and
%$0113$ and the second, labeled $01.2$ is constructed by combining
%$0121,\ldots 01.24$. Finally, in $A_2$ all groups are combined.
%
%To compute an aggregate for records labeled with $0111$, the algorithm will
%select those records and test whether the subset is suitable. If not, it will
%move to $f_1(0111)=011$ and find all the groups in $A_0$ that map to 011 as
%follows.
%\begin{displaymath}
%f_1^{-1}(\{011\}) = \{0111, 0112, 0113\}.
%\end{displaymath}
%It will then move on and find all records labeled with any of those three
%group labels, and test again. If this new data set is suitable, the result is
%computed, if not, another collapse takes place. The whole procedure is repeated 
%for each element in $A_0$.

Denote with $F_k:A_0\to A_k$, the function that  accepts a label in $A_0$ and
returns the corresponding label in $A_k$. This can formally be defined as
consecutive application of $f_1, f_2,\cdots, f_k$ to an element of $A_0$. In
notation
\begin{equation*}
F_k \equiv f_k\circ f_{k-1}\circ\cdots\circ f_1.
\end{equation*}
Similarly we define $F^{-1}_k:2^{A_k}\to 2^{A_0}$ as the composition
\begin{equation*}
F_k^{-1} \equiv f_1^{-1}\circ f_2^{-1}\circ\cdots\circ f_k^{-1}.
\end{equation*}
This function asks for a set of labels in $A_k$ and returns all the labels in
$A_0$ that are mapped to those labels via the collapsing sequence of
Equation~\eqref{eq:collapsingsequence}.  We are now ready to formally write
down our main Algorithm~\ref{alg:saccg}.

%
\begin{algorithm}[H]
\caption{Split-Apply-Combine with Collapsing Groups}
\label{alg:saccg}
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
\Input{A finite set $U$, an aggregator $\phi: 2^U\to X$, a test function $\beta: 2^U\to \mathbb{B}$,
      and a collapsing sequence $U\xrightarrow{f_0}A_0\cdots \xrightarrow{f_n}  A_n$}

@@

$R = \{\}$\;
\For{$a\in A_0$}{
  $i=0$\;
  $d = f_0^{-1}(\{a\})$\tcp*{get subset of $U$}
  \While{$i<n \land \lnot\beta(d)$}{
    $i = i+1$ \tcp*{Increase collapse level} 
    $d = (f_0^{-1}\circ F_i^{-1}\circ F_i)(a)$ \tcp*{Collapse and get subset}
  }
  \If{$i<n \lor \beta(d)$}{ \label{line:cond}
    $R = R\cup \{(a, i,\phi(d))\}$\;
  } 
}

\end{algorithm}
%
Since the used level of collapsing is determined dynamically by data
circumstances in $U$, the algorithm also reports the collapsing level used to
compute a value foe each member of $A_0$. For each member of $A_0$ there is a
triple $(a,i,\phi(d))$, where $a\in A_0$ is the label to which the value
$\phi(d)$ pertains, and $i$ is the number of collapses applied to reach a
suitable dataset. The condition in Line~\ref{line:cond} ensures that if no
suitable dataset is found after the whole collapsing sequence has been
executed, then no answer is returned.

>>>>

% shortcut for surjection arrow, and surjection version of xrightarrow
\newcommand{\onto}{\twoheadrightarrow}
\newcommand\xonto[2][]{%
  \mathrel{\ooalign{$\xrightarrow[#1\mkern4mu]{#2\mkern4mu}$\cr%
  \hidewidth$\rightarrow\mkern4mu$}}
}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Formal description and algorithms}
In this section we give a formal description of the algorithm for aggregation
with dynamic grouping.  We start by giving an algorithm for ordinary
split-apply-combine to show where the algorithm must be generalized to allow
for a collapsing scheme.

\subsection{Split-apply-combine}
To analyse a data set group by group we need to specify a data set, a
way to split it into groups, and a function that takes a subset of data and
returns an aggregate. Let us introduce some notation for that.

Denote with $U$ be a finite set, and and let $\phi:2^U\to X$ be a function that
accepts a subset of $U$ and returns a value in some domain $X$.  Here, $U$
represents a data set, $2^U$ its power set, and $\phi$ an aggregating function.
We split $U$ into groups using the following notation. Let $A$ be finite set
that has no more elements than $U$, and let $f:\onto A$ be a
surjective function that takes an element of $U$ and returns a value in $A$. We
can think of $A$ as a set of group labels, and $f$ as the function that assigns
a label to each element of $U$. This way, $f$ divides $U$ into non-overlapping
subsets. We say that $f:U\onto A$ is a \emph{partition} of $U$. We
also introduce the \emph{pullback along $f$}, $f^*:2^A\to 2^U$ defined as
%
\begin{equation*}
f^*(S) = \{u\in U|f(u)\in S\},
\end{equation*}
where $S$ is a subset of $A$ \citep[Section 1.4]{fong2019invitation}.

@@

\begin{algorithm}[H]
\caption{Split-Apply-Combine}
\label{alg:sac}
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
\Input{A finite set $U$, an aggregator $\phi: 2^U\to X$, and a partition
       $f:U\onto A$.}
\Output{$R$: the value of $\phi$ for every part of $U$ as a set of pairs
        $(a,x)\in A\times X$. }
$R = \{\}$\;
\For{$a\in A$}{
  $d=f^*(\{a\})$\tcp*{get subset of $U$}
  $R = R\cup \{(a,\phi(d))\}$\tcp*{aggregate and add to result}
}
\end{algorithm}
%
In this algorithm the output is collected in a set $R$ containing pairs from
$A\times X$: one pair for each element of $A$. Incidentely, the algorithm can
be summarized even shorter in this notation as $R=\cup_{a\in A}(a,(\phi\circ
f^*)(\{a\}))$, where $\circ$ denotes function composition.  Since $f^*$ always
receives a singleton set, its definition as a function of the powerset of the
group labels $A$ is too general for this algorithm.  However, this
generalization turns out to be necessary when generalizing the algorithm to the
case of dynamic grouping.

It is interesting to see how the elements $U$, $f:U\onto A$, and
$\phi$ are implemented in practice. Consider the signature of the
\proglang{R}'s \code{aggregate()} function (we skip arguments that are not
important for the discussion).
\begin{verbatim}
  aggregate(x, by, FUN)
\end{verbatim}
Here, \code{x} is a data frame where each column is a variable $U$ to be
aggregated.  The parameter \code{by} is a list of vectors of group labels,
where each vector has a with a length that equals the number of rows in
\code{x}. So the function $f:U\onto A$ is implemented by asking
the user to make sure that the position of each label corresponds to the
correct row number in the data frame \code{x}. The argument \code{FUN} is the
function $\phi$ that aggregates each subset of \code{x}.  Here is an example of
how a user might call this function from the R prompt.
%
<<>>=
aggregate(iris[1:2], by=iris["Species"], FUN=mean)
@
%
Note that the correspondence in position of the \texttt{Species} label and the
record position is implemented by taking them from the same data frame. The
output also reveals in the first column the set $A$: each row corresponds to a
unique value in \texttt{Species} column.

\subsection{Split-apply-combine with collapsing groups}
\label{sect:saccg}
The goal of the algorithm is to compute a value for each part of a dataset,
possibly using values external to the part. The input of the algorithm consists
again of a finite set $U$ and an aggregation function $\phi$ that takes a
subset of $U$ and returns a value in some domain $X$.  Compared to
Algorithm~\ref{alg:sac} two different inputs are needed. First, a function must
be defined that that checks whether a given subset $d$ of $U$ is suitable for
computing $\phi(d)$.  We will call this function $\beta: 2^U\to \mathbb{B}$,
where $\mathbb{B}=\{\texttt{True},\texttt{False}\}$.  Typical tests are
checking whether there are sufficient records available, or whether certain
variables have a low enough fraction of missing values. Second, we need a
\emph{collapsing scheme}
\begin{equation}
U\xonto{f_0}A_0\xonto{f_1}A_1\xonto{f_2}\cdots\xonto{f_n}A_n.
\label{eq:collapsingsequence}
\end{equation}
A collapsing scheme is a sequence of partitions where each $f_i:A_{i-1}\to A_i$
partitions $A_{i-1}$ into $|A_i|$ groups.

Denote with $F_k:A_0\to A_k$, the function that  accepts a label in $A_0$ and
returns the corresponding label in $A_k$. In other words, $F_k$ is the
composition $f_k\circ f_{k-1}\circ\cdots \circ f_1$. Similarly we define the
pullback along $F_k$ as $F_k^* = f_1^*\circ f_2^*\circ\cdots\circ f_k^*$.  This
function accepts a set of labels in $A_k$ and returns all the labels in $A_0$
that are mapped to those labels via the collapsing sequence of
Equation~\eqref{eq:collapsingsequence}.  With this notation we can define the
Algorithm for aggregation with dynamic grouping as follows.

\begin{algorithm}[H]
\caption{Split-Apply-Combine with Collapsing Groups}
\label{alg:saccg}
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
\Input{A finite set $U$, an aggregator $\phi: 2^U\to X$, a test function $\beta: 2^U\to \mathbb{B}$,
      and a collapsing sequence $U\xonto{f_0}A_0\xonto{f_1}\cdots \xonto{f_n}  A_n$}

@@

$R = \{\}$\;
\For{$a\in A_0$}{
  $i=0$ \tcp*{Initiate collapse level}
  $d = f_0^*(\{a\})$ \tcp*{Get subset of $U$}
  \While{$i<n \land \lnot\beta(d)$}{ \label{line:while}
    $i = i+1$ \tcp*{Increase collapse level} 
    $d = (f_0^*\circ F_i^*\circ F_i)(a)$ \tcp*{Collapse and get subset}
  }\label{line:endwhile}
  \If{$i<n \lor \beta(d)$}{ \label{line:cond}
    $R = R\cup \{(a, i,\phi(d))\}$\;
  } 
}

\end{algorithm}
%

In this algorithm the collapsing level $i$ is increased until the test is
passed or the maximum collapsing level $n$ is reached
(Lines~\ref{line:while}-\ref{line:endwhile}) %Since the eventual level of
collapsing is determined dynamically by data %circumstances in $U$, the
algorithm also reports the collapsing level $i$ used.  The condition in
Line~\ref{line:cond} ensures that if no suitable dataset is found after the
whole collapsing sequence has been executed, then no answer is returned. This
means that in contrast with Algorithm~\ref{alg:sac} there is no guarantee that
a value for each member of $A_0$ will be found. For each member of $A_0$ where
an aggregate is computed, there is a triple $(a,i,\phi(d))$, where $a\in A_0$
is the label to which the value $\phi(d)$ pertains, and $i$ is the number of
collapses applied to reach a suitable dataset.
